\section{Simpler forms of Soundness}

\subsection{Environment Narrowing and Transitivity}
A central difficulty in achieving a proof of soundness is the interplay between environment narrowing (the modification of an environment to a more specific one) and subtype transitivity. The proofs of soundness for DOT deal with this by pushing back on transitivity using \emph{tight} and \emph{inert} typing.
In this paper we demonstrate that a much simplified type system can easily achieve soundness while still containing a variety of interesting features. Primarily we remove environment narrowing altogether by using a double headed subtyping relation ($\Gamma_1 \vdash \tau_1 <: \tau_2 \dashv \Gamma_2$). \citeauthor{nada 2014} used a similar subtyping relation when developing soundness using a big step semantics. We remove environment narrowing in the absence of intersection types.

\subsection{Path Dependent Types}

DOT only allows type selections on variables ($x$). This disqualifies the use of more complex type selections that include field access in the path ($x.f_1 \ldots .f_n$). Such paths are required for modelling modules as objects. They hold some complexity in the evaluation of the paths. Consider the following example.
\begin{lstlisting}[mathescape, style=custom_lang]
val x = new {
        def m(x : {z $\Rightarrow$
                   type L = $\top$
          }) = { x } : x.L
        }
\end{lstlisting}
In this case we define a method that takes a parameter \verb|x| and returns a term of type \verb|x.L|. Now if we attempt to type the following code in this context.
\begin{lstlisting}[mathescape, style=custom_lang]
val z = new {type L = $\top$}
val y = new {
         val f = z
         }
x.m(y)
\end{lstlisting}
\verb|x.m(y.f)| has type \verb|y.f.L|. Further reduction eventually results in \verb|z.L|. This is not the same type and so violates the type preservation portion of type safety.

We aim to introduce complex paths as part of selection types. The strategy involved here is to include paths as values, that is paths are not evaluated.
