\section{A Class-Based Language}

This version adds classes and shows how to rewrite them in terms of
more primitive constructs.  It is a true subset of the real Wyvern
language.

Note that \{\} are used in the abstract syntax as an abbreviation for
an indented block.


\input{f-core3-syntax}


\clearpage


\begin{figure}
\centering
\[
\begin{array}{c}
\infer[\textit{DT-type}]
	{\Gamma, \_ \vdash \keyw{type}~ t ~ \{\overline{\tau_d}\} :: \keyw{type}~ t ~ \{\overline{\tau_d}\} } 
	{}\\[3ex]
\infer[\textit{DT-class}]
	{\Gamma, \_ \vdash \keyw{class}~ c ~ \{\overline{cd}, \overline d\} :: \keyw{class}~ c ~ \{\overline{cd}, \overline d\} } 
	{\Gamma, \sigma_{this} \vdash \overline d :: \overline{\sigma_d} & \Gamma, \sigma_{this} \vdash \overline{cd} :: \overline{\sigma_{cd}} & \sigma_{this} = \{\overline{\sigma_{cd}}, \overline{\sigma_d} \} }\\[3ex]
\infer[\textit{DT-class-var}]
	{\Gamma, \sigma_{this} \vdash \keyw{class var} ~ f : \tau_1 = e_1 :: \keyw{class var}~ f : \tau_1 } 
	{\Gamma, \sigma_{this} \vdash e : \tau_2 & \tau_2 <: \tau_1 }\\[3ex]
\infer[\textit{DT-class-meth}]
	{\Gamma, \sigma_{this} \vdash \keyw{class meth} ~ m : \tau_1 = e_1 :: \keyw{class meth}~ m : \tau_1 } 
	{\Gamma, \sigma_{this} \vdash e : \tau_2 & \tau_2 <: \tau_1 }\\[3ex]
\infer[\textit{T-new}]
	{\Gamma, \sigma_{this} \vdash \keyw{new}~ \{ \overline{d_1} \} : \{ \overline{\tau_d} \}}
	{\Gamma,\sigma_{this} \vdash \overline{d_1} :: \overline{\sigma_{d_1}} & \{\overline{\sigma_{d_1}}\} \union \sigma_{this} <: \{\overline{\tau_d} \} } \\[3ex]
\infer[\textit{T-class-field}]
	{\Gamma, \sigma_{this} \vdash  e.f:\tau_1} 
	{\Gamma, \sigma_{this} \vdash e : \sigma_1 & \sigma_1 = \{\keyw{class var}~ f:\tau_1,... \} }\\[3ex]
\infer[\textit{T-class-meth}]
	{\Gamma, \sigma_{this} \vdash  e.m:\tau_1} 
	{\Gamma, \sigma_{this} \vdash e : \tau_1 & \tau_1 = \{\keyw{class meth}~ m:\tau_1=e_1,... \} }
\end{array}
\]
\caption{Static Semantics Rules Core 3}
\end{figure}


\input{f-core3-translate-function}


\clearpage


\subsection{Example Program and Translation by Darya}

\input{f-core3-example}
\input{f-core3-example-translated}


\clearpage


\subsection{Example Program in the Class-Based Language (By Jonathan)}

\input{f-core3-example-old}

\subsection{Translation of the Program to the Core Method-Based
  Language (By Jonathan)}

\input{f-core3-example-translated-old}

Limitations: this language only supports objects, not ADTs.  For ADTs we need bounded type members.


\clearpage


\subsection{Tasks}

\begin{itemize}

 \item write some examples!
 \item define stripClass, rewriteNew, and a way of computing $\tau_i$
 \item add lots of conveniences as sugar
 \item in rule R-class, meth c needs to return the same object each time, so cache it in a field.
 \item give complete rewriting rules (R*) to the core language
 \item give complete typing rules, and prove that well-typed source programs translate to well-typed core programs.  Is it possible to prove a property related to the uniform access principle and/or state encapsulation?
 \item consider ``class type t = ...''
 \item no abstract class members
 \item no class class members because a class is always a class member; if you want a class in an object use a val $+$ type 

\end{itemize}
